The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor-Couette flow

### 2017-03-02 paper/thingap_ref_response/thingap_mri_paper_ref_response_apj.tex
new paper draft

\documentclass{emulateapj}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\hyd}    {{\rm H}}
\newcommand{\vobs}{v_{\rm obs}}
\newcommand{\hi}{H{\sc i}~}
\newcommand{\hii}{H{\sc ii}~}
\newcommand{\hia}{H{\sc i}}
\newcommand{\hiia}{H{\sc ii}}

\usepackage[colorlinks,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{gensymb}
\usepackage{amsmath}
\usepackage{color}

\usepackage[titletoc,title]{appendix}

\makeatletter
\setlength{\@fptop}{0pt}
\setlength{\@fpbot}{0pt}
\setlength{\@fpsep}{0pt}
\makeatother

\newcommand{\citei}[1]{\citeauthor{#1} \citeyear{#1}}
\newcommand{\unit}[1]{\textrm{ #1}}
\newcommand{\kms}{km ${\rm s^{-1}}$}
\newcommand{\kmsa}{km ${\rm s^{-1}}$}
\newcommand{\citeia}[2]{\citeauthor{#1}~(\citeyear{#1};~#2)}

@@

\renewcommand{\topfraction}{0.85}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.75}

@@

\title{The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor-Couette flow}
\author{S. E. Clark\altaffilmark{1}}
\author{Jeffrey S. Oishi\altaffilmark{2,}\altaffilmark{3}}

\altaffiltext{1}{Department of Astronomy, Columbia University, New York, NY} 
\altaffiltext{2}{Department of Physics and Astronomy, Bates College, Lewiston, ME}
\altaffiltext{3}{Department of Astrophysics, American Museum of Natural History, New York, NY}

\begin{abstract}
The magnetorotational instability (MRI) is a fundamental process of accretion disk physics, but its saturation mechanism remains poorly understood despite considerable theoretical and computational effort. We present a multiple scales analysis of the non-ideal MRI in the weakly nonlinear regime -- that is, when the most unstable MRI mode has a growth rate asymptotically approaching zero from above. Here, we develop our theory in a thin-gap, Cartesian channel. Our results confirm the finding by Umurhan et al. (2007) that the perturbation amplitude follows a Ginzburg-Landau equation. We extend these results by performing a detailed force balance for the saturated azimuthal velocity and vertical magnetic field, demonstrating that even when diffusive effects are important, the bulk flow saturates via the combined processes of reducing the background shear and rearranging and strengthening the background vertical magnetic field. We directly simulate the Ginzburg-Landau amplitude evolution for our system and demonstrate the pattern formation our model predicts on long length and time scales. 
\end{abstract}

\section{Introduction}

For matter to accrete from a disk onto a central object, angular momentum must be transported radially outward in the disk. The transport mechanism is likely turbulent, as molecular viscosity alone cannot account for the needed angular momentum transfer, and likely magnetic, as this turbulence is excited even in hydrodynamically stable disks (\citei{Shakura:1973wg}). Discovered by \citet{Chandrasekhar:1960wh} and \citet{Velikhov:1959} in a global geometry, the magnetorotational instability (MRI) was subsequently rediscovered and applied to accretion disks by \citet{Balbus:1991vs}. Since then, the MRI remains the leading explanation for rapid angular momentum transport in astrophysical disks. The instability in its simplest geometry arises when a differentially rotating disk is threaded by a vertical magnetic field. The presence of the magnetic field linearly destabilizes the disk gas, driving turbulence and angular momentum transport \citep[e.g.][]{2011ApJ...738...84H,2014MNRAS.438.2513P,2013MNRAS.435.2281P}. The MRI likely plays a role in a diverse host of astrophysical systems, including protoplanetary disks \citep[e.g.][]{2015ApJ...798...84B} and black hole accretion disks \citep[e.g.][]{2013ApJ...769..156S}, as well as stellar interiors \citep[e.g.][]{2015ApJ...799...85W}. Despite its importance, many aspects of the MRI remain poorly understood. In particular, the nonlinear saturation mechanism for the MRI is an open question, and a formidable challenge. MRI saturation has been tackled almost exclusively with simulation, with a few notable exceptions detailed below. In this work we analytically investigate the weakly nonlinear saturation of the MRI.

Weakly nonlinear analysis is a perturbative method used to examine the asymptotic behavior of a system near threshold -- that is, when the system is just barely unstable to its most unstable mode. The analytical technique follows the multiscale evolution of fluid variables in a perturbation expansion, allowing the controlled interaction of modes between orders in a perturbation series \citep{1978amms.book.....B}. Weakly nonlinear analysis can be a powerful technique for analytically examining systems which in their full generality exhibit such complicated nonlinear behavior that their study is relegated primarily to the simulation domain. The MRI is one such phenomenon: while there is a rich literature analytically examining the linear MRI, analytical treatments of the \textit{nonlinear} system are relatively few. The weakly nonlinear treatment of the MRI was pioneered by \citet{Knobloch:2005ba} and \citet[][hereafter URM07; see also \citeyear{Umurhan:2007dz}]{Umurhan:2007hs}. The latter authors undertook the first weakly nonlinear analysis of the MRI in a thin-gap Taylor-Couette (TC) flow with strong dissipation (as is appropriate to liquid metal experiments), and found that the marginal MRI system approaches saturation in a manner analogous to that of Rayleigh-B\'enard convection. Weakly nonlinear analysis was instrumental in our understanding of Rayleigh-B\'enard convection saturation (\citei{Newell:1969}), and the similarities between convection and the thin-gap MRI are the result of important shared symmetries between the systems. The success of URM07 in modeling the MRI system near threshold merits further consideration, but we are unaware of any other attempts to expand upon their theoretical framework. In this work we rederive the theory of URM07, and expand upon their findings. Our focus here is on fully characterizing the thin-gap MRI system, both for independent theoretical interest and to have a robust comparison point for extensions of this theory into more complicated geometries. In a companion paper we derive for the first time the weakly nonlinear theory of the standard and helical MRI in a global, cylindrical TC flow \citep[][hereafter Paper II]{Clark:2016}. The thin- and wide-gap treatments complement one another theoretically, and both are important regimes for comparison with simulation.

@@

We note several other important analytical studies of MRI saturation. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They consider a thin-gap regime as well, and so their theory may be considered the strongly nonlinear analogue to the one developed here. \citei{Vasil:2015} examines the weakly nonlinear MRI in a thin-gap regime in a minimal model, finding deep mathematical similarities between the MRI system and the elastodynamic instability of a buckling beam. We discuss these results and their relation to ours in section~\ref{sec:discussion}. Several authors have investigated the behavior of the MRI when the boundary conditions are shear periodic, and so the MRI has no mechanism by which to modify the background shear flow profile. In this approximation linear MRI growth is dominated by channel modes, a type of MRI mode that, for periodic boundary conditions, are exact solutions of both the linear and nonlinear MRI equations (\citei{Goodman:1994ul}). In this regime the MRI saturates via parasitic instabilities, which feed off and destroy the primary MRI modes. Analytical investigation of this case reveals that MRI saturation can be caused by parasitic Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime (\citei{Pessah:2010ic}). The theory of MRI channel mode parasites is robust (see also \citei{Pessah:2009gm}, \citei{Latter:2010iz}, \citei{Rembiasz:2016}), but their importance may be overestimated by the local approximation \citep{Latter:2015}, and not germane to global analyses like the one presented here. \citet{Latter:2015} gives a detailed analysis of the relation between local and global linear MRI modes.

@@

\beq\label{eq:momentum}
\begin{split}
& \partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, = \, -\frac{1}{\rho}\nabla P \, - \, \nabla\Phi \, + \, \frac{1}{\rho} \left(\mathbf{J}\times\mathbf{B}\right) \, \\
& + \, \nu\nabla^2 \mathbf{u} \, - \, 2\mathbf{\Omega} \times \mathbf{u} \, - \, \mathbf{\Omega} \times \left(\mathbf{\Omega} \times \mathbf{r} \right), \\
\end{split}
\eeq

@@

\begin{multline}
\label{eqset1}
\partial_t \nabla^2 \Psi \, + \, J\left(\Psi, \nabla^2 \Psi\right) \, - \, 2 \partial_z u_{y} \, = \\
\, \frac{2}{\beta} B_0 \partial_z \nabla^2 A \, + \, \frac{2}{\beta}J\left(A, \nabla^2 A \right) \, + \, \frac{1}{\reye}\nabla^4 \Psi
\end{multline}

\begin{multline}
\label{eqset2}
\partial_t u_{y} \, + \, J\left(\Psi, u_{y}\right) \, + \, \left(2 - q\right) \Omega_0 \partial_z \Psi \, = \\
\, \frac{2}{\beta}B_0\partial_z B_{y} \, + \, \frac{2}{\beta} J\left(A, B_{y}\right) \, + \, \frac{1}{\reye} \nabla^2 u_{y}
\end{multline}

\begin{multline}
\label{eqset3}
\partial_t A \, = \, B_0 \partial_z \Psi \, + \, J\left(A, \Psi\right) \, + \, \frac{1}{\reym} \nabla^2 A
\end{multline}

\begin{multline}
\label{eqset4}
\partial_t B_{y} \, = \, B_0 \partial_z u_{y} \, + \, J\left(A, u_{y}\right) \, - \, J\left(\Psi, B_{y}\right) \, \\
+ \, \frac{1}{\reym} \nabla^2 B_{y}  \, - \, q \Omega_0 \partial_z A,
\end{multline}

@@

\begin{align}
\mathcal{O}(\epsilon)&: \mathcal{L}\mathbf{V_1} + \mathcal{D}\partial_t \mathbf{V_1} = 0. \label{eq:ordere}\\
\mathcal{O}(\epsilon^2)&: \mathcal{L}\mathbf{V_2} + \widetilde{\mathcal{L}}_1 \partial_Z \mathbf{V_1} = \mathbf{N_2} \label{eq:ordere2}\\
\mathcal{O}(\epsilon^3)&: \mathcal{L} \mathbf{V_3} + \mathcal{D}\partial_T \mathbf{V_1}  + \widetilde{\mathcal{L}}_1\partial_Z\mathbf{V_2} + \widetilde{\mathcal{L}}_2\partial_Z^2\mathbf{V_1} + \widetilde{\mathcal{G}}\mathbf{V_1} = \mathbf{N_3} \label{eq:ordere3}
\end{align}

The partial differential equations that comprise Equations \ref{eq:ordere} to \ref{eq:ordere3} are solved in succession. The practical advantage of our ansatz construction (Equation \ref{V1_ansatz}) is clear: the separable x-dependence means that the radial boundary conditions are solved in only one dimension. Thus our analytical framework is able to side-step many of the resolution issues faced by multidimensional simulations. We are able to resolve even small-scale structure in the boundary layers of our domain, because we need only resolve it in one dimension. We solve the radial component of each equation using the open source pseudospectral code Dedalus. We compute the radial components on a grid of Chebyshev polynomials, as is appropriate for bounded one-dimensional domains (e.g. \citei{Boyd:2001aa}). The nonuniform spacing of the Chebyshev grid allows us to resolve the boundary layers well on a 128-point grid.

@@

\subsection{\textbf{Shearing box and ambipolar diffusion}}

@@

@@

The fact that we find a Ginzburg-Landau equation for the amplitude is unsurprising: such an amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}. \textbf{The geometry of the problem considered here mimics an idealized experimental setup, where the local MRI equations are subject to wall-like radial boundary conditions.} In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well. \textbf{How would the theory developed here change with the addition of vertical endcaps, as found in a real experiment?}%\textbf{Many studies of the MRI take place in the local shearing box, in which Equations \ref{eqset1} - \ref{eqset4} are solved subject to shear-periodic boundary conditions. This is meant }

@@

\clearpage
\appendix

\section{A. Detailed Equations}\label{app:matrices}

Here we detail the perturbation analysis described in Section \ref{sec:wnl}. The perturbation series is described by Equations \ref{eq:ordere} - \ref{eq:ordere3}, where

\beq
\mathcal{L} = \mathcal{L}_0 + \mathcal{L}_1 \partial_z + \mathcal{L}_2 \partial_z^2 + \mathcal{L}_3 \partial_z^3 + \mathcal{L}_4 \partial_z^4,
\eeq

\beq
\widetilde{\mathcal{L}}_1 =  \mathcal{L}_1 + 2\mathcal{L}_2\partial_z + 3\mathcal{L}_3\partial_z^2 + 4\mathcal{L}_4\partial_z^3
\eeq

\beq
\widetilde{\mathcal{L}}_2 = \mathcal{L}_2 + 3\mathcal{L}_3\partial_z + 6\mathcal{L}_4\partial_z^2
\eeq

\beq
\widetilde{\mathcal{G}} = - \mathcal{G} \partial_z - \mathcal{L}_3 \partial_z^3,
\eeq

and the constituent matrices are defined as

\beq
\mathcal{D} = \left[\begin{matrix}
\nabla^2 & 0 & 0 & 0 \\
0 & 1& 0 & 0 \\
0 & 0 & 1 & 0\\
0 & 0 & 0 & 1 \\
\end{matrix}\right]
\eeq

>>>>

%%
%% Beginning of file 'sample61.tex'
%%
%% Modified 2016 September
%%
%% This is a sample manuscript marked up using the
%% AASTeX v6.1 LaTeX 2e macros.
%%
%% AASTeX is now based on Alexey Vikhlinin's emulateapj.cls 
%% (Copyright 2000-2015).  See the classfile for details.

%% AASTeX requires revtex4-1.cls (http://publish.aps.org/revtex4/) and
%% other external packages (latexsym, graphicx, amssymb, longtable, and epsf).
%% All of these external packages should already be present in the modern TeX 
%% distributions.  If not they can also be obtained at www.ctan.org.

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style  which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%%
%% using aastex version 6.1
\documentclass[twocolumn]{aastex61}

\usepackage{amsmath}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}

@@

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be envoked like this:
%%
%% \documentclass[argument]{aastex61}
%% 
%% where the arguement options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can envoke to allow other stylistic
%% actions. The available options are:
%%
%%  astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%  tighten      : Makes baselineskip slightly smaller, only works with 
%%                 the twocolumn substyle.
%%  times        : uses times font instead of the default
%%  linenumbers  : turn on lineno package.
%%  trackchanges : required to see the revision mark up and print its output
%%  longauthor   : Do not use the more compressed footnote style (default) for 
%%                 the author/collaboration/affiliations. Instead print all
%%                 affiliation information after each name. Creates a much
%%                 long author list but may be desirable for short author papers
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex61}

%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%%\hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003

%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
%\received{July 1, 2016}
%\revised{September 27, 2016}
%\accepted{\today}
%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the arguement.
\submitjournal{ApJ}

%% Mark up commands to limit the number of authors on the front page.
%% Note that in AASTeX v6.1 a \collaboration call (see below) counts as
%% an author in this case.
%
%\AuthorCollaborationLimit=3
%
%% Will only show Schwarz, Muench and "the AAS Journals Data Scientist 
%% collaboration" on the front page of this example manuscript.
%%
%% Note that all of the author will be shown in the published article.
%% This feature is meant to be used prior to acceptance to make the
%% front end of a long author article more manageable. Please do not use
%% this functionality for manuscripts with less than 20 authors. Conversely,
%% please do use this when the number of authors exceeds 40.
%%
%% Use \allauthors at the manuscript end to show the full author list.
%% This command should only be used with \AuthorCollaborationLimit is used.

%% The following command can be used to set the latex table counters.  It
%% is needed in this document because it uses a mix of latex tabular and
%% AASTeX deluxetables.  In general it should not be needed.
%\setcounter{table}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{The weakly nonlinear magnetorotational instability}
\shortauthors{Clark \& Oishi}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%  \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% This is the end of the preamble.  Indicate the beginning of the
%% manuscript itself with \begin{document}.

@@

\title{The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor-Couette flow}

%% LaTeX will automatically break titles if they run longer than
%% one line. However, you may use \\ to force a line break if
%% you desire. In v6.1 you can include a footnote in the title.

%% A significant change from earlier AASTEX versions is in the structure for 
%% calling author and affilations. The change was necessary to implement 
%% autoindexing of affilations which prior was a manual process that could 
%% easily be tedious in large author manuscripts.
%%
%% The \author command is the same as before except it now takes an optional
%% arguement which is the 16 digit ORCID. The syntax is:
%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
%%
%% This will hyperlink the author name to the author's ORCID page. Note that
%% during compilation, LaTeX will do some limited checking of the format of
%% the ID to make sure it is valid.
%%
%% Use \affiliation for affiliation information. The old \affil is now aliased
%% to \affiliation. AASTeX v6.1 will automatically index these in the header.
%% When a duplicate is found its index will be the same as its previous entry.
%%
%% Note that \altaffilmark and \altaffiltext have been removed and thus 
%% can not be used to document secondary affiliations. If they are used latex
%% will issue a specific error message and quit. Please use multiple 
%% \affiliation calls for to document more than one affiliation.
%%
%% The new \altaffiliation can be used to indicate some secondary information
%% such as fellowships. This command produces a non-numeric footnote that is
%% set away from the numeric \affiliation footnotes.  NOTE that if an
%% \altaffiliation command is used it must come BEFORE the \affiliation call,
%% right after the \author command, in order to place the footnotes in
%% the proper location.
%%
%% Use \email to set provide email addresses. Each \email will appear on its
%% own line so you can put multiple email address in one \email call. A new
%% \correspondingauthor command is available in V6.1 to identify the
%% corresponding author of the manuscript. It is the author's responsibility
%% to make sure this name is also in the author list.
%%
%% While authors can be grouped inside the same \author and \affiliation
%% commands it is better to have a single author for each. This allows for
%% one to exploit all the new benefits and should make book-keeping easier.
%%
%% If done correctly the peer review system will be able to
%% automatically put the author and affiliation information from the manuscript
%% and save the corresponding author the trouble of entering it by hand.

\correspondingauthor{S. E. Clark}
\email{seclark@astro.columbia.edu}

\author[0000-0002-7633-3376]{S. E. Clark}
\affil{Department of Astronomy, Columbia University, New York, NY}

\author{Jeffrey S. Oishi}
\affiliation{Department of Physics and Astronomy, Bates College, Lewiston, ME}
\affiliation{Department of Astrophysics, American Museum of Natural History, New York, NY}
\nocollaboration

%% Note that the \and command from previous versions of AASTeX is now
%% depreciated in this version as it is no longer necessary. AASTeX 
%% automatically takes care of all commas and "and"s between authors names.

%% AASTeX 6.1 has the new \collaboration and \nocollaboration commands to
%% provide the collaboration status of a group of authors. These commands 
%% can be used either before or after the list of corresponding authors. The
%% argument for \collaboration is the collaboration identifier. Authors are
%% encouraged to surround collaboration identifiers with ()s. The 
%% \nocollaboration command takes no argument and exists to indicate that
%% the nearby authors are not part of surrounding collaborations.

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}

The magnetorotational instability (MRI) is a fundamental process of accretion disk physics, but its saturation mechanism remains poorly understood despite considerable theoretical and computational effort. We present a multiple scales analysis of the non-ideal MRI in the weakly nonlinear regime -- that is, when the most unstable MRI mode has a growth rate asymptotically approaching zero from above. Here, we develop our theory in a thin-gap, Cartesian channel. Our results confirm the finding by Umurhan et al. (2007) that the perturbation amplitude follows a Ginzburg-Landau equation. We extend these results by performing a detailed force balance for the saturated azimuthal velocity and vertical magnetic field, demonstrating that even when diffusive effects are important, the bulk flow saturates via the combined processes of reducing the background shear and rearranging and strengthening the background vertical magnetic field. We directly simulate the Ginzburg-Landau amplitude evolution for our system and demonstrate the pattern formation our model predicts on long length and time scales.

\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% See the online documentation for the full list of available subject
%% keywords and the rules for their use.
%\keywords{editorials, notices --- miscellaneous --- catalogs --- surveys}

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.

%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

\section{Introduction} \label{sec:intro}
For matter to accrete from a disk onto a central object, angular momentum must be transported radially outward in the disk. The transport mechanism is likely turbulent, as molecular viscosity alone cannot account for the needed angular momentum transfer, and likely magnetic, as this turbulence is excited even in hydrodynamically stable disks \citep{Shakura:1973wg}. Discovered by \citet{Chandrasekhar:1960wh} and \citet{Velikhov:1959} in a global geometry, the magnetorotational instability (MRI) was subsequently rediscovered and applied to accretion disks by \citet{Balbus:1991vs}. Since then, the MRI remains the leading explanation for rapid angular momentum transport in astrophysical disks. The instability in its simplest geometry arises when a differentially rotating disk is threaded by a vertical magnetic field. The presence of the magnetic field linearly destabilizes the disk gas, driving turbulence and angular momentum transport \citep[e.g.][]{2011ApJ...738...84H,2014MNRAS.438.2513P,2013MNRAS.435.2281P}. The MRI likely plays a role in a diverse host of astrophysical systems, including protoplanetary disks \citep[e.g.][]{2015ApJ...798...84B} and black hole accretion disks \citep[e.g.][]{2013ApJ...769..156S}, as well as stellar interiors \citep[e.g.][]{2015ApJ...799...85W}. Despite its importance, many aspects of the MRI remain poorly understood. In particular, the nonlinear saturation mechanism for the MRI is an open question, and a formidable challenge. MRI saturation has been tackled almost exclusively with simulation, with a few notable exceptions detailed below. In this work we analytically investigate the weakly nonlinear saturation of the MRI.

Weakly nonlinear analysis is a perturbative method used to examine the asymptotic behavior of a system near threshold -- that is, when the system is just barely unstable to its most unstable mode. The analytical technique follows the multiscale evolution of fluid variables in a perturbation expansion, allowing the controlled interaction of modes between orders in a perturbation series \citep{1978amms.book.....B}. Weakly nonlinear analysis can be a powerful technique for analytically examining systems which in their full generality exhibit such complicated nonlinear behavior that their study is relegated primarily to the simulation domain. The MRI is one such phenomenon: while there is a rich literature analytically examining the linear MRI, analytical treatments of the \textit{nonlinear} system are relatively few. The weakly nonlinear treatment of the MRI was pioneered by \citet{Knobloch:2005ba} and \citet[][hereafter URM07; see also \citeyear{Umurhan:2007dz}]{Umurhan:2007hs}. The latter authors undertook the first weakly nonlinear analysis of the MRI in a thin-gap Taylor-Couette (TC) flow with strong dissipation (as is appropriate to liquid metal experiments), and found that the marginal MRI system approaches saturation in a manner analogous to that of Rayleigh-B\'enard convection. Weakly nonlinear analysis was instrumental in our understanding of Rayleigh-B\'enard convection saturation \citep{Newell:1969}, and the similarities between convection and the thin-gap MRI are the result of important shared symmetries between the systems. The success of URM07 in modeling the MRI system near threshold merits further consideration, but we are unaware of any other attempts to expand upon their theoretical framework. In this work we rederive the theory of URM07, and expand upon their findings. Our focus here is on fully characterizing the thin-gap MRI system, both for independent theoretical interest and to have a robust comparison point for extensions of this theory into more complicated geometries. In a companion paper we derive for the first time the weakly nonlinear theory of the standard and helical MRI in a global, cylindrical TC flow \citep[][hereafter Paper II]{Clark:2016}. The thin- and wide-gap treatments complement one another theoretically, and both are important regimes for comparison with simulation.

@@

We note several other important analytical studies of MRI saturation. \citeauthor{Knobloch:2005ba} (\citeyear{Knobloch:2005ba}) analyze the MRI in the strongly nonlinear regime, by following the already-developed MRI modes into asymptotic saturation. They consider a thin-gap regime as well, and so their theory may be considered the strongly nonlinear analogue to the one developed here. \citet{Vasil:2015} examines the weakly nonlinear MRI in a thin-gap regime in a minimal model, finding deep mathematical similarities between the MRI system and the elastodynamic instability of a buckling beam. We discuss these results and their relation to ours in section~\ref{sec:discussion}. Several authors have investigated the behavior of the MRI when the boundary conditions are shear periodic, and so the MRI has no mechanism by which to modify the background shear flow profile. In this approximation linear MRI growth is dominated by channel modes, a type of MRI mode that, for periodic boundary conditions, are exact solutions of both the linear and nonlinear MRI equations \citep{Goodman:1994ul}. In this regime the MRI saturates via parasitic instabilities, which feed off and destroy the primary MRI modes. Analytical investigation of this case reveals that MRI saturation can be caused by parasitic Kelvin-Helmholtz and tearing mode instabilities, depending on parameter regime \citep{Pessah:2010ic}. The theory of MRI channel mode parasites is robust \citep[see also]{Pessah:2009gm;Latter:2010iz;Rembiasz:2016}, but their importance may be overestimated by the local approximation \citep{Latter:2015}, and not germane to global analyses like the one presented here. \citet{Latter:2015} gives a detailed analysis of the relation between local and global linear MRI modes.

@@

\begin{eqnarray}\label{eq:momentum}
\begin{split}
& \partial_t \mathbf{u} \, + \, \mathbf{u} \cdot \nabla \mathbf{u} \, = \, -\frac{1}{\rho}\nabla P \, - \, \nabla\Phi \, + \, \frac{1}{\rho} \left(\mathbf{J}\times\mathbf{B}\right) \, \\
& + \, \nu\nabla^2 \mathbf{u} \, - \, 2\mathbf{\Omega} \times \mathbf{u} \, - \, \mathbf{\Omega} \times \left(\mathbf{\Omega} \times \mathbf{r} \right), \\
\end{split}
\end{eqnarray}

@@

%\begin{widetext}
\onecolumngrid
\beq
\label{eqset1}
\partial_t \nabla^2 \Psi \, - \, 2 \partial_z u_{y} \, - \, \frac{2}{\beta} B_0 \partial_z \nabla^2 A \, - \, \frac{1}{\reye}\nabla^4 \Psi \, = \, \frac{2}{\beta}J\left(A, \nabla^2 A \right) \, - \, J\left(\Psi, \nabla^2 \Psi\right) \, 
\eeq
%\end{eqnarray}
%\end{widetext}

\beq
\label{eqset2}
\partial_t u_{y} \, + \, \left(2 - q\right) \Omega_0 \partial_z \Psi \, = \, \frac{2}{\beta}B_0\partial_z B_{y} \, + \, \frac{2}{\beta} J\left(A, B_{y}\right) \, - \, J\left(\Psi, u_{y}\right) \, + \, \frac{1}{\reye} \nabla^2 u_{y}
\eeq

\beq
\label{eqset3}
\partial_t A \, = \, B_0 \partial_z \Psi \, + \, J\left(A, \Psi\right) \, + \, \frac{1}{\reym} \nabla^2 A
\eeq

\beq
\label{eqset4}
\partial_t B_{y} \, = \, B_0 \partial_z u_{y} \, + \, J\left(A, u_{y}\right) \, - \, J\left(\Psi, B_{y}\right) \, \\
+ \, \frac{1}{\reym} \nabla^2 B_{y}  \, - \, q \Omega_0 \partial_z A,
\eeq
\twocolumngrid

@@

\begin{eqnarray}
\mathcal{O}(\epsilon)&: \mathcal{L}\mathbf{V_1} + \mathcal{D}\partial_t \mathbf{V_1} = 0. \label{eq:ordere}\\
\mathcal{O}(\epsilon^2)&: \mathcal{L}\mathbf{V_2} + \widetilde{\mathcal{L}}_1 \partial_Z \mathbf{V_1} = \mathbf{N_2} \label{eq:ordere2}\\
\mathcal{O}(\epsilon^3)&: \mathcal{L} \mathbf{V_3} + \mathcal{D}\partial_T \mathbf{V_1}  + \widetilde{\mathcal{L}}_1\partial_Z\mathbf{V_2} + \widetilde{\mathcal{L}}_2\partial_Z^2\mathbf{V_1} + \widetilde{\mathcal{G}}\mathbf{V_1} = \mathbf{N_3} \label{eq:ordere3}
\end{eqnarray}

The partial differential equations that comprise Equations \ref{eq:ordere} to \ref{eq:ordere3} are solved in succession. The practical advantage of our ansatz construction (Equation \ref{V1_ansatz}) is clear: the separable x-dependence means that the radial boundary conditions are solved in only one dimension. Thus our analytical framework is able to side-step many of the resolution issues faced by multidimensional simulations. We are able to resolve even small-scale structure in the boundary layers of our domain, because we need only resolve it in one dimension. We solve the radial component of each equation using the open source pseudospectral code Dedalus. We compute the radial components on a grid of Chebyshev polynomials, as is appropriate for bounded one-dimensional domains \citep[e.g.]{Boyd:2001aa}. The nonuniform spacing of the Chebyshev grid allows us to resolve the boundary layers well on a 128-point grid.

@@

\subsection{\edit1{Shearing box and ambipolar diffusion}}

@@

The Ginzburg-Landau form of the amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}. In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well.
The Ginzburg-Landau equation arises due to symmetries in the local MRI equations, irrespective of the boundary conditions to which they are subjected. This means that the local MRI is able to saturate via nonlinear mode interaction so long as the primary MRI modes are not exact solutions of the nonlinear terms. This can be achieved by considering the effects of ambipolar diffusion when the boundary conditions are shear periodic, or by enforcing wall-like radial boundary conditions. Both constructions require the most unstable mode to have nonconstant radial structure. %So long as our system permits the most unstable mode to nonlinearly interact with

%The fact that we find a Ginzburg-Landau equation for the amplitude is unsurprising: such an amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}. In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well. \textbf{How would the theory developed here change with the addition of vertical endcaps, as found in a real experiment?}%\textbf{Many studies of the MRI take place in the local shearing box, in which Equations \ref{eqset1} - \ref{eqset4} are solved subject to shear-periodic boundary conditions. This is meant }

@@

%The fact that we find a Ginzburg-Landau equation for the amplitude is unsurprising: such an amplitude equation can be found in any system with Euclidean symmetry and a quadratic maximum in growth rate with respect to the wavenumber \citep{Hoyle:2006}. In this case, the Euclidean symmetry comes from axisymmetry in the $x$-$z$ plane, and the quadratic maximum is a consequence of the linear dispersion relation given in Appendix~\ref{app:dispersion}. In Paper II, we show that the same symmetry occurs in the axisymmetric global geometry as well. \textbf{How would the theory developed here change with the addition of vertical endcaps, as found in a real experiment?}%\textbf{Many studies of the MRI take place in the local shearing box, in which Equations \ref{eqset1} - \ref{eqset4} are solved subject to shear-periodic boundary conditions. This is meant }

@@

\appendix
\begin{thebibliography}{}
